---
title: "Polling Places Analysis"
author: "Justin S."
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
code_folding: TRUE
---

In this analysis, I will look at Federal Sentencing data from the [United States Sentencing Commission (USSC)](https://www.ussc.gov/). I will use a variety of exploratory and modeling techniques to do answer the following questions:


# Let's Talk About the Data
The data provided by USSC has proved challenging in a rather unique way. The datasets are saved as zip files, one for each fiscal year, containing .SAS and .SPSS files. Unfortunately, those files are fixed delimited, but not totally consistently, making it hard to actually put the data into a standard, tabular format, despite the fact that that was what was intended by the USSC research team compiling the data. 

With that, huge kudos is in order to Kevin Wilson for writing [significant code](https://github.com/khwilson/SentencingCommissionDatasets) to convert each zip file into a CSV. Even after the couple hours spent running his code, I was able to produce one CSV for each fiscal years' data. Unfortunately, each of the 11 CSVs was about 1GB...which is what you get when you have 100,000 rows and 400 variables in each file, so I had to compress the files to better work with them. Even then, having 1mil+ rows and 400 columns wasn't too feasible, so I spent significant time going through [USSC's Codebook](https://www.ussc.gov/sites/default/files/pdf/research-and-publications/datafiles/USSC_Public_Release_Codebook_FY99_FY20.pdf) to better understand which variables they provide. With that, my analysis will be tailored to answer the following questions:

```{r fig.align='center', fig.height=6, fig.width=6, message=FALSE, warning=FALSE, include=FALSE}
# Set our plot specifications for the rest of the document.
knitr::opts_chunk$set(fig.width = 9,
                      fig.height = 6,
                      fig.align = "center",
                      # Set our code specifications for the rest of the document
                      echo = F,
                      warning = F,
                      message = F)
```

Let's take a look at the data. Here's a small sample of it:

```{r set-up_and_clean_data, include=FALSE}
###########################################################################
## Set Up -----------------------------------------------------------------
###########################################################################
# Bring in packages
suppressMessages(library("pacman"))
pacman::p_load("tidyverse", # Used for data wrangling,
               "tidyr", # Used for data cleaning,
               "ggplot2", # Used for visualizations,
               "ggrepel", # Used for labeling points in a ggplot viz
               "here", # Used for navigating project structure
               "maps", # Used for map-based visualizations
               "readxl", # Used for loading excel files,
               "readr", # Used for working with files,
               "pander", # Used for pretty tables,
               "kableExtra", # Used for RMarkdown formatting
               "lubridate", # Used for fixing dates,
               "usmap", # Used for plotting US states
               "praise", # Used for positive reinforcement,
               "janitor", # Used for data cleaning,
               "pdftools", # Used for reading PDF files in,
               "gganimate", # Used for interactive graphic visualizations,
               "gridExtra", # Used for putting multiple plots next to each other
               "mapproj", # Used for visualizing maps
               "transformr", # Used to animate maps
               "gifski", # Used to create animated gifs
               "forecast", # Used for time series analysis,
               "tseries")  # Used for time series analysis

# Get rid of scientific notation
options(scipen = 999)

# Try unzipping the files
files <- unzip('Data/Processed/Archive.zip')

### Bring in the data, taking advantage of the project structure
# Note: There are 11 CSVs, one for each year between 2010-2020,
#       so we'll bring them all in programmatically into one df
# Initialize an empty tibble to hold all of our data
sentencing_data <- dplyr::tibble()

for (csv in unzip('Data/Processed/Archive.zip', list = TRUE)$Name) {
  # Where we at
  print(paste0("Reading in ", csv, "..."))
  
  # Read it in as a temporary tibble
  temp_data <- read_csv(unzip('Data/Processed/Archive.zip', csv),
    col_names = TRUE,
    # Only pull in columns we care about to speed up the process
    # Definitions and descriptions of all variables are available in the codebook here:
    # https://www.ussc.gov/sites/default/files/pdf/research-and-publications/datafiles/USSC_Public_Release_Codebook_FY99_FY20.pdf
    col_types = cols_only(
      AGE = col_double(), # Actual Age
      YEARS = col_double(), # Age Category buckets
      ALTDUM = col_logical(), # Dummy indicator of alternative sentence
      ALTMO = col_double(), # Total months of alternative incarceration
      AMENDYR = col_character(), # The guideline manual used to calculate the guideline calculations
      AMTFINEC = col_double(), # Total dollar amount of both fine and cost of supervision
      AMTREST = col_double(), # Dollar amount of restitution
      FINE = col_double(), # Dollar amount of fine ordered (including cost of supervision)
      FINEWAIV = col_logical(), # Indicates if fine was waived because of defendant's inability to pay
      AMTTOTAL = col_double(), # Sum of the imposed dollar amounts of fine, cost of supervision, and restitution
      ARMCRIM = col_factor(), # # of lvls to be subtracted due to defendant's acceptance of responsibility
      CAROFFAP = col_logical(), # Indicator to whether career offender status was applied
      CAROFFEN = col_factor(), # # of lvls to subtract from the Career Offender Offense Level due to defendant's Acceptance of Responsibility
      CARROFFLV = col_double(), # Offense level due to the application of career offender status
      CASETYPE = col_factor(), # 1 = Felony; 2 = Misdemenaor A; 3 = Misdemeanor B/C (2018-present)
      CIRCDIST = col_double(), # Districts in the order in which they appear in Sourcebook
      CITIZEN = col_factor(), # 1 = US Citizen; 2 = Resident; 3 = Undocumented Immigrant; 4 = Unknown; 5 = Extradited Alien
      NEWCIT = col_logical(), # 0 = U.S. Citizen; 1 = Non-U.S. Citizen
      CITWHERE = col_character(), # Defendant's country of citizenship
      COMDUM = col_logical(), # Indicator whether defendant received community confinement
      MOCOMCON = col_double(), # Total months of community confinement
      COSTSUP = col_double(), # Cost of supervision, only if ordered separately from fine
      CRIMHIST = col_logical(), # Whether defendant has ANY criminal history or LE contacts
      CRIMLIV = col_double(),
      CRIMPTS = col_double(),
      CRMLIVAP = col_factor(),
      CRPTS = col_logical(),
      DAYSDUM = col_logical(), # Indicator whether offender received a sentence of any days imposed in prison
      DEFCONSL = col_factor(), # Type of defense counsel used in the instant offense
      DISPOSIT = col_factor(), # Disposition of the defendant's case
      DOB = col_date(), # Date of Birth
      DOBMON = col_factor(), # Month of Birth
      DOBYR = col_double(), # Year of Birth
      DRUGMIN = col_double(), # Mandatory min sentence (in months) associated with drug statutes
      'DRUGTYP1' = col_character(), # Drug types involved in a case
      'DRUGTYPX' = col_character(), # Drug types involved in a case
      DTGDL = col_logical(), # Identifies cases with one of the "Big 7" drug guidelines as the primary guideline
      ECONDUM = col_logical(), # Indicator of whether a fine/cost of supervision or restitution amount was given
      EDUCATN = col_factor(), # Highest lvl of education completed by defendant
      HISPORIG = col_factor(), # 0 = No Info available; 1 = Non-Hispanic; 2 = Hispanic
      HOMDUM = col_logical(), # Indicates whether defendant received home detention
      MOHOMDET = col_double(), # Total months of home detention ordered
      HRCOMSRV = col_double(), # Total # of comm service hours ordered
      INOUT = col_logical(), # 0 = Received Prison Sentence; 1 = Received Non-Prison Sentence
      INTDUM = col_logical(), # Indicates whether a defendant received intermittent confinement
      MOINTCON = col_double(), # Total months of intermittent confinement ordered
      'LOSS1' = col_double(), # Dollar amount of loss for which the offender is responsible
      'LOSSX' = col_double(), # Same as above
      'LOSSHI' = col_double(), # Same as above
      MAND1 = col_factor(), # Status of any mandatory minimums at sentencing
      MANDX = col_factor(), # Status of any mandatory minimums at sentencing
      MARRIED = col_factor(), # Marital status
      MNTHDEPT = col_double(), # Diff in months between guideline minimum (GLMIN) and sentence length (SENSPCAP)
      MONCIRC = col_factor(), # Judicial circuit in which the defendant was sentenced
      MONRACE = col_factor(), # Offender's race
      MONSEX = col_logical() # FALSE/0 = Male; TRUE/1 = Female
    ),
    # n_max = 10000,
    progress = show_progress()
    ) %>% 
    janitor::remove_empty(which = c("rows", "cols")) %>% 
    as_tibble()
  
  sentencing_data <- bind_rows(temp_data, sentencing_data)
}

```


Let's start by taking a look at the data.

```{r display_data}
########################################################################
### View Data --------------------------------------------------------
########################################################################
# Let's view the data output
# In any kable outputs, display NAs as blanks
opts <- options(knitr.kable.NA = "")

sentencing_data %>% 
  head(25) %>%
  # Fix up the headers by replacing the underscores with spaces
  rename_all(funs(str_replace_all(., "_", " "))) %>% 
  # Make everything proper capitalization
  # rename_all(funs(str_to_title)) %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = F,
                font_size = 10) %>%
  # Make the header row bold and black so it's easier to read
  row_spec(0, bold = T, color = "black") %>% 
  scroll_box(height = "400px", width = "100%")
```
